---
title: "Lake Ice Duration vs. Winter Air Temperature"
author: "Joachim Vogl"
date: "`r Sys.Date()`"
output:
  github_document: default
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/",
  fig.width = 7,
  fig.height = 5
)
```

## Research Question

Is mean winter air temperature significantly related to mean annual lake ice
duration, and can it be used to predict ice duration on lakes?

## Data

Both datasets come from the North Temperate Lakes Long Term Ecological Research
site (Madison, Wisconsin) via the `lterdatasampler` package:

- `ntl_icecover` — annual ice cover duration (days) for Lake Mendota and Lake Monona
- `ntl_airtemp` — daily average air temperature (°C) for Madison, WI

```{r packages}
library(tidyverse)
library(lterdatasampler)
library(rstatix)

data("ntl_icecover")
data("ntl_airtemp")
```

## 1. Build the Analysis Table

Ice duration is averaged across the two lakes for each water year (October –
September), and daily air temperature is averaged over the winter months
(November – March) of each water year. The two tables are then joined on water
year.

```{r build-table}
# mean ice duration per water year (water year = the following calendar year)
avg_icecover <- ntl_icecover %>%
  group_by(wyear = year + 1) %>%
  summarize(mean_duration = mean(ice_duration, na.rm = TRUE))

# assign each daily air temperature reading to a water year (Oct-Dec roll forward)
ntl_airtemp_wyear <- ntl_airtemp %>%
  mutate(wyear = if_else(month(sampledate) < 10, year, year + 1))

# mean winter (Nov-Mar) air temperature per water year
ntl_airtemp_winter_avg <- ntl_airtemp_wyear %>%
  filter(month(sampledate) %in% c(11, 12, 1, 2, 3)) %>%
  group_by(wyear) %>%
  summarize(winter_average_temp = mean(ave_air_temp_adjusted, na.rm = TRUE))

# join winter temperature to ice duration
icecover_temp <- inner_join(ntl_airtemp_winter_avg, avg_icecover, by = "wyear")

head(icecover_temp)
```

## 2. Explore the Data

```{r theme}
# shared look for every plot
my_theme <- theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"),
        axis.text  = element_text(color = "black"))
```

### Distributions

```{r hist-air-temp}
ggplot(icecover_temp, aes(x = winter_average_temp)) +
  geom_histogram(binwidth = 0.5, fill = "orange", color = "white") +
  labs(title = "Distribution of Average Winter Air Temp",
       x = "Average Winter Air Temp (°C)",
       y = "Number of Winters") +
  my_theme
```

```{r hist-ice-duration}
ggplot(icecover_temp, aes(x = mean_duration)) +
  geom_histogram(binwidth = 5, fill = "orange", color = "white") +
  labs(title = "Distribution of Mean Ice Cover Duration",
       x = "Mean Ice Cover Duration (days)",
       y = "Number of Winters") +
  my_theme
```

Both variables are roughly bell-shaped and unimodal. Ice duration has a small
tail of unusually short ice seasons (under ~60 days) and one very long one
(~160 days), so a rank-based Spearman correlation is used below.

### Trends Over Time

```{r trend-air-temp}
ggplot(icecover_temp, aes(x = wyear, y = winter_average_temp)) +
  geom_point(colour = "orange") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "purple", fill = "purple", alpha = 0.2) +
  labs(title = "Average Air Temp by Winter Year",
       x = "Winter Year",
       y = "Average Air Temp (°C)") +
  my_theme
```

```{r trend-ice-duration}
ggplot(icecover_temp, aes(x = wyear, y = mean_duration)) +
  geom_point(colour = "orange") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "purple", fill = "purple", alpha = 0.2) +
  labs(title = "Mean Ice Cover Duration by Winter Year",
       x = "Winter Year",
       y = "Mean Ice Cover Duration (days)") +
  my_theme
```

### Ice Duration vs. Winter Air Temperature

```{r ice-vs-temp}
ggplot(icecover_temp, aes(x = winter_average_temp, y = mean_duration)) +
  geom_point(colour = "orange") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "purple", fill = "purple", alpha = 0.2) +
  labs(title = "Mean Ice Cover Duration vs. Average Winter Air Temp",
       x = "Average Winter Air Temp (°C)",
       y = "Mean Ice Cover Duration (days)") +
  my_theme
```

The relationship is clearly negative and approximately linear.

## 3. Correlation Test

```{r correlation}
# Spearman = non-parametric (rank-based) correlation
cor_test(icecover_temp,
         vars = c(mean_duration, winter_average_temp),
         alternative = "two.sided",
         method = "spearman")
```

There is a significant negative relationship between mean ice duration and mean
winter air temperature — warmer winters have shorter ice cover. The Spearman
correlation coefficient is **-0.84** (p = 2.67e-41, so p < 0.05), a strong
negative correlation.

## 4. Simple Linear Regression

```{r slr-model}
duration_temp_model <- lm(mean_duration ~ winter_average_temp, data = icecover_temp)

summary(duration_temp_model)
```

- **Slope:** -8.112 days of ice per °C
- **Intercept:** 71.735 days of ice (predicted duration at 0 °C)
- **Residual standard error:** 11.01 days on 150 degrees of freedom
- **R²:** 0.660

## 5. Predictions

Predicted mean ice duration when mean winter air temperature is -2 °C, 0 °C,
and 2 °C:

```{r predictions}
new_durationtemp <- tibble(winter_average_temp = c(-2, 0, 2))

predict(duration_temp_model, newdata = new_durationtemp)
```

| Mean winter air temp | Predicted mean ice duration |
|---------------------:|----------------------------:|
| -2 °C                | 87.96 days                  |
| 0 °C                 | 71.74 days                  |
| 2 °C                 | 55.51 days                  |
