---
title: "Palmer Penguins: Species Summary Statistics & Distributions"
author: "Joachim Vogl"
date: "`r Sys.Date()`"
output:
  github_document: default
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/",
  fig.width = 7,
  fig.height = 5,
  dpi = 150,
  fig.retina = 1
)
```

## Research Question

How do Adélie, Chinstrap, and Gentoo penguins differ in body size, and what do
the distributions of their bill and flipper measurements look like?

## Data

The `penguins` dataset from the `palmerpenguins` package contains size
measurements for 344 adult penguins of three species, collected on three islands
in the Palmer Archipelago, Antarctica, from 2007 to 2009 by Dr. Kristen Gorman and
the Palmer Station Long Term Ecological Research (LTER) program. The data were
originally published in Gorman, Williams & Fraser (2014), *PLoS ONE* 9(3): e90081.

```{r packages}
library(tidyverse)
library(palmerpenguins)
library(knitr)

glimpse(penguins)
```

```{r theme}
# shared look for every plot
my_theme <- theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"),
        axis.text  = element_text(color = "black"),
        legend.title = element_text(face = "bold"))

species_colors <- c(Adelie = "darkorange", Chinstrap = "purple", Gentoo = "cyan4")
```

## 1. Sample Size by Species

```{r species-counts}
table(penguins$species)   # the three counts separately
nrow(penguins)            # total number of penguins
```

## 2. Bill Length

Average bill length for all penguins, then by species in millimeters and inches
(1 mm = 0.03937008 in).

```{r bill-length}
mm_to_in <- 0.03937008

# all penguins
mean(penguins$bill_length_mm, na.rm = TRUE)
mean(penguins$bill_length_mm, na.rm = TRUE) * mm_to_in

# by species
penguins %>%
  group_by(species) %>%
  summarize(mean_bill_mm = mean(bill_length_mm, na.rm = TRUE)) %>%
  mutate(mean_bill_in = mean_bill_mm * mm_to_in) %>%
  kable(digits = 3,
        col.names = c("Species", "Mean bill length (mm)", "Mean bill length (in)"))
```

## 3. Descriptive Statistics by Species

Sample size, percent male, and flipper length statistics for each species and
for the full sample.

```{r summary-table}
summarize_group <- function(df) {
  df %>%
    summarize(
      n               = n(),
      pct_male        = 100 * sum(sex == "male", na.rm = TRUE) / n(),
      mean_flipper_mm = mean(flipper_length_mm, na.rm = TRUE),
      sd_flipper_mm   = sd(flipper_length_mm, na.rm = TRUE),
      max_flipper_mm  = max(flipper_length_mm, na.rm = TRUE)
    )
}

species_stats <- penguins %>%
  group_by(species) %>%
  summarize_group() %>%
  mutate(species = as.character(species))

overall_stats <- penguins %>%
  summarize_group() %>%
  mutate(species = "All penguins")

bind_rows(species_stats, overall_stats) %>%
  kable(digits = 1,
        col.names = c("Species", "n", "% Male", "Mean flipper (mm)",
                      "SD flipper (mm)", "Max flipper (mm)"))
```

*Percent male is out of all penguins in each group; sex was not recorded for 11
penguins (6 Adélie, 5 Gentoo).*

## 4. Distribution of Bill Depth

```{r hist-bill-depth}
ggplot(penguins, aes(x = bill_depth_mm, fill = species)) +
  geom_histogram(binwidth = 0.5, color = "white") +
  scale_fill_manual(values = species_colors) +
  labs(title = "Distribution of Penguin Bill Depth",
       x = "Bill Depth (mm)",
       y = "Number of Penguins",
       fill = "Species") +
  my_theme
```

```{r bill-depth-stats}
penguins %>%
  group_by(species) %>%
  summarize(mean_depth_mm = mean(bill_depth_mm, na.rm = TRUE),
            sd_depth_mm   = sd(bill_depth_mm, na.rm = TRUE)) %>%
  kable(digits = 2,
        col.names = c("Species", "Mean bill depth (mm)", "SD bill depth (mm)"))
```

Bill depth is bimodal overall: Gentoo penguins have much shallower bills than
Adélie and Chinstrap penguins, which overlap almost completely.

## 5. Relationships Between Measurements

### Flipper Length vs. Bill Depth

```{r flipper-vs-depth}
ggplot(penguins, aes(x = flipper_length_mm, y = bill_depth_mm,
                     color = species, shape = species)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = species_colors) +
  labs(title = "Flipper Length vs. Bill Depth by Species",
       x = "Flipper Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species", shape = "Species") +
  my_theme
```

### Bill Length vs. Bill Depth

```{r bill-length-vs-depth}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm,
                     color = species, shape = species)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = species_colors) +
  labs(title = "Bill Length vs. Bill Depth by Species",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species", shape = "Species") +
  my_theme
```

```{r bill-correlations}
# correlation across all penguins
cor(penguins$bill_length_mm, penguins$bill_depth_mm, use = "complete.obs")

# correlation within each species
penguins %>%
  group_by(species) %>%
  summarize(r = cor(bill_length_mm, bill_depth_mm, use = "complete.obs")) %>%
  kable(digits = 2, col.names = c("Species", "Correlation (r)"))
```

Across all penguins, bill length and bill depth are weakly *negatively*
correlated, but within every species the relationship is *positive*. Pooling the
species hides the real pattern (an example of Simpson's paradox).

### Bill Dimensions by Sex, Split by Species

```{r bill-by-sex}
penguins %>%
  filter(!is.na(sex)) %>%
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = sex, shape = sex)) +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~ species) +
  scale_color_manual(values = c(female = "#C2185B", male = "#1565C0")) +
  labs(title = "Bill Length vs. Bill Depth by Sex",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Sex", shape = "Sex") +
  my_theme +
  theme(strip.text = element_text(face = "bold"))
```

Within every species, males tend to have both longer and deeper bills than
females.
